---
title: "A Brief data transformation and visualization"
author: "Kawin Chinpong"
date: "`r Sys.Date()`"
output: html_document
---

This is a R markdown which were demonstrated some (maybe) useful information on how tabular transformation and basic visualization is.

# Outline on this markdown

Information on this markdown is consisted of the following sections

- Library Loading
- (incidence table) Data Loading
- Data conversion from wide-format into long-format
- Basic visualization

# Library Loading

In this chapter, `tidyverse` which also included `readr` for data preprocessing and `ggplot2` for visualization, and `rgdal` will be loaded for main data performing while `sf` is for map transformation. Library `DT` were only used for tabular visualization in R-markdown only, not necessary.

```{r}
library(tidyverse)
library(rgdal)
library(sf)
library(DT)
```

# (incidence table) Data Loading

Suppose we quarterly collected incidence count of a disease for 2 consecutive years. Original result data maybe gathered like the following table, read it using `read_csv` which were loaded from `readr`.

```{r}
df_wide<-read_csv("input/incidence_table_wide.csv")
datatable(df_wide,caption="Quarter Incidence Table in Wide-format",options=list(pageLength=5,scrollX=TRUE))
```

Unfortunately, appropriate format should be like this following table.

```{r}
df_long<-read_csv("input/incidence_table_long.csv")
datatable(df_long,caption="Quarter Incidence Table in Long-format",options=list(pageLength=5,scrollX=TRUE))
```

Therefore, data conversion should be early applied before visualization.

# Data conversion

Original table in wide-format can be transformed into long-format using `pivot_longer` loaded from `tidyr`. Essential arguments in this function is detailed as follow

- `cols`: specific columns that should be melted, specify as vector of string, or pattern-based function such as `starts_with`, `ends_with`, `contains`.
- `names_to`: speficy column name after melted down columns specified in `cols`
- `values_to`: specify column name of values within columns specified in `cols` after melting process.

```{r}
df<-df_wide %>% pivot_longer(cols=c("q1","q2","q3","q4"),names_to = 'num_quarter', values_to = 'incidence_count')
# df<-df %>% mutate(num_quarter=as.integer(str_remove_all(num_quarter,'q')))
datatable(df,caption="Quarter Incidence Table in appropriate format",options=list(pageLength=5,scrollX=TRUE))
```

# Basic Visualization

In this section, `df` which were appropriate format incidence data will be plot as Thailand map. Try to load shapefile retrieved from GADM.

```{r}
tha_map <- readOGR("input/gadm41_THA_shp",layer="gadm41_THA_1")
datatable(tha_map@data,caption="Data retrieve from GADM",options=list(pageLength=5,scrollX=TRUE))
```

Next, try to merge map with incidence data with `sp::merge` and then convert to `sf` format

```{r}
tha_map_merge <- sp::merge(tha_map,df,by.x="NAME_1",by.y="area",duplicateGeoms=TRUE)
tha_map_merge_sf <- st_as_sf(tha_map_merge)
```

Finally, try to plot using `ggplot` which simple settings were shown as afterward snippets.

```{r}
ggplot(tha_map_merge_sf) + geom_sf(aes(fill=incidence_count)) + facet_wrap(vars(year,num_quarter),dir="h",ncol=4)
```

and here is another customization.

```{r}
ggplot(tha_map_merge_sf) + geom_sf(aes(fill=incidence_count)) + facet_wrap(vars(year,num_quarter),dir="h",ncol=4)+ggtitle("Quarter Incidence Count")+theme_bw()+theme(axis.text.x=element_blank(),axis.text.y=element_blank(),axis.ticks=element_blank())+scale_fill_gradient2(low='green',mid='yellow',high='red',midpoint=1500)
```

Moreover, above visualization could be exported as image file.

```{r}
ggsave("output/result.png")
```

# Some references

- https://readr.tidyverse.org/
- https://tidyr.tidyverse.org/
- https://cran.r-project.org/web/packages/rgdal/index.html
- https://ggplot2.tidyverse.org/